---
title: "MC_scATAC"
output: html_document
date: "2022-12-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ArchR)
library(BSgenome.Hsapiens.UCSC.hg19)
```


# Create Arrow files
```{r}
addArchRThreads(threads = 8) 
```
```{r}
inputFiles <- getTutorialData("Hematopoiesis")
inputFiles
```

```{r}
addArchRGenome("hg19")
```

```{r}
ArrowFiles <- createArrowFiles(
  inputFiles = inputFiles,
  sampleNames = names(inputFiles),
  filterTSS = 4, #Dont set this too high because you can always increase later
  filterFrags = 1000, 
  addTileMat = TRUE,
  addGeneScoreMat = TRUE
)
```

# Doublets detection

```{r}
doubScores <- addDoubletScores(
    input = ArrowFiles,
    k = 10, #Refers to how many cells near a "pseudo-doublet" to count.
    knnMethod = "UMAP", #Refers to the embedding to use for nearest neighbor search with doublet projection.
    LSIMethod = 1
)
```

# Create ArchRProject
```{r}
projHeme1 <- ArchRProject(
  ArrowFiles = ArrowFiles, 
  outputDirectory = "HemeOutput",
  copyArrows = TRUE
)
```

```{r}
getAvailableMatrices(projHeme1)
```

```{r}
head(projHeme1$cellNames)
```

```{r}
table(projHeme1$Sample)
```

```{r}
quantile(projHeme1$TSSEnrichment)
```


```{r}
QC_df <- getCellColData(projHeme1, select = c("log10(nFrags)", "TSSEnrichment"))
QC_df
```

```{r}
class(QC_df)
```

# Data QC 

```{r}
ggPoint(
    x = QC_df[,1], 
    y = QC_df[,2], 
    colorDensity = TRUE,
    continuousSet = "sambaNight",
    xlabel = "Log10 Unique Fragments",
    ylabel = "TSS Enrichment",
    xlim = c(log10(500), quantile(QC_df[,1], probs = 0.99)),
    ylim = c(0, quantile(QC_df[,2], probs = 0.99))
) + geom_hline(yintercept = 4, lty = "dashed") + geom_vline(xintercept = 3, lty = "dashed")
```

## QC for each sample
```{r}
p1=plotGroups(
    ArchRProj = projHeme1, 
    groupBy = "Sample", 
    colorBy = "cellColData", 
    name = "TSSEnrichment",
    plotAs = "violin",
    alpha = 0.4,
    addBoxPlot = TRUE
   )
p2=plotGroups(
    ArchRProj = projHeme1, 
    groupBy = "Sample", 
    colorBy = "cellColData", 
    name = "log10(nFrags)",
    plotAs = "violin",
    alpha = 0.4,
    addBoxPlot = TRUE
   )
p1+p2
```

```{r}
p1 <- plotFragmentSizes(ArchRProj = projHeme1)
p2 <- plotTSSEnrichment(ArchRProj = projHeme1)
p1+p2
```
```{r}
saveArchRProject(ArchRProj = projHeme1, outputDirectory = "Save-ProjHeme1", load = FALSE)
```

## Filter Doublets
```{r}
projHeme2 <- filterDoublets(projHeme1)

```

# Dimension reduction

```{r}
projHeme2 <- addIterativeLSI(
    ArchRProj = projHeme2,
    useMatrix = "TileMatrix", 
    name = "IterativeLSI", 
    iterations = 2, 
    clusterParams = list( #See Seurat::FindClusters
        resolution = c(0.2), 
        sampleCells = 10000, 
        n.start = 10
    ), 
    varFeatures = 25000, 
    dimsToUse = 1:30
)
```

```{r}
projHeme2 <- addHarmony(
    ArchRProj = projHeme2,
    reducedDims = "IterativeLSI",
    name = "Harmony",
    groupBy = "Sample"
)
```

# Clustering

```{r}
projHeme2 <- addClusters(
    input = projHeme2,
    reducedDims = "IterativeLSI",
    method = "Seurat",
    name = "Clusters",
    resolution = 0.8
)
```

```{r}
table(projHeme2$Clusters)
```

```{r}
projHeme2 <- addClusters(
    input = projHeme2,
    reducedDims = "Harmony",
    method = "Seurat",
    name = "HarmonyClusters",
    resolution = 0.8
)
```

```{r}
table(projHeme2$HarmonyClusters)
```

# Embeddings
```{r}
projHeme2 <- addUMAP(
    ArchRProj = projHeme2, 
    reducedDims = "IterativeLSI", 
    name = "UMAP", 
    nNeighbors = 30, 
    minDist = 0.5, 
    metric = "cosine"
)
```

```{r}
projHeme2 <- addUMAP(
    ArchRProj = projHeme2, 
    reducedDims = "Harmony", 
    name = "UMAPHarmony", 
    nNeighbors = 30, 
    minDist = 0.5, 
    metric = "cosine"
)
```

```{r}
p1 <- plotEmbedding(ArchRProj = projHeme2, colorBy = "cellColData", name = "Sample", embedding = "UMAP")
p2 <- plotEmbedding(ArchRProj = projHeme2, colorBy = "cellColData", name = "Clusters", embedding = "UMAP")
p1+p2
```

```{r}
p3 <- plotEmbedding(ArchRProj = projHeme2, colorBy = "cellColData", name = "Sample", embedding = "UMAPHarmony")
p4 <- plotEmbedding(ArchRProj = projHeme2, colorBy = "cellColData", name = "Clusters", embedding = "UMAPHarmony")
p3+p4
```

```{r}
p5 <- plotEmbedding(ArchRProj = projHeme2, colorBy = "cellColData", name = "Sample", embedding = "UMAPHarmony")
p6 <- plotEmbedding(ArchRProj = projHeme2, colorBy = "cellColData", name = "HarmonyClusters", embedding = "UMAPHarmony")
p5+p6
```
# Identify marker genes

```{r}
markersGS <- getMarkerFeatures(
    ArchRProj = projHeme2, 
    useMatrix = "GeneScoreMatrix", 
    groupBy = "Clusters",
    bias = c("TSSEnrichment", "log10(nFrags)"),
    testMethod = "wilcoxon"
)
```

```{r}
markerList <- getMarkers(markersGS, cutOff = "FDR <= 0.01 & Log2FC >= 1.25")

```

```{r}
markerList$C12
```

```{r}
markerGenes  <- c(
    "CD34", #Early Progenitor
    "GATA1", #Erythroid
    "PAX5", "MS4A1", "EBF1", "MME", #B-Cell Trajectory
    "CD14", "CEBPB", "MPO", #Monocytes
    "IRF8", 
    "CD3D", "CD8A", "TBX21", "IL7R" #TCells
  )

heatmapGS <- markerHeatmap(
  seMarker = markersGS, 
  cutOff = "FDR <= 0.01 & Log2FC >= 1.25", 
  labelMarkers = markerGenes,
  transpose = TRUE
)
```

```{r}
ComplexHeatmap::draw(heatmapGS, heatmap_legend_side = "bot", annotation_legend_side = "bot")

```
```{r}
markerGenes  <- c(
    "CD34",  #Early Progenitor
    "GATA1", #Erythroid
    "PAX5", "MS4A1", "MME", #B-Cell Trajectory
    "CD14", "MPO", #Monocytes
    "CD3D", "CD8A"#TCells
  )

p <- plotEmbedding(
    ArchRProj = projHeme2, 
    colorBy = "GeneScoreMatrix", 
    name = markerGenes, 
    embedding = "UMAP",
    quantCut = c(0.01, 0.95),
    imputeWeights = NULL
)

p2 <- lapply(p, function(x){
    x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
        axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()
    )
})
do.call(cowplot::plot_grid, c(list(ncol = 3),p2))
```
# Impute gene scores
```{r}
projHeme2 <- addImputeWeights(projHeme2)

```

```{r}
p <- plotEmbedding(
    ArchRProj = projHeme2, 
    colorBy = "GeneScoreMatrix", 
    name = markerGenes, 
    embedding = "UMAP",
    imputeWeights = getImputeWeights(projHeme2)
)

#Rearrange for grid plotting
p2 <- lapply(p, function(x){
    x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
        axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()
    )
})
do.call(cowplot::plot_grid, c(list(ncol = 3),p2))
```
```{r}
p <- plotBrowserTrack(
    ArchRProj = projHeme2, 
    groupBy = "Clusters", 
    geneSymbol = "CD14", 
    upstream = 50000,
    downstream = 50000
)

grid::grid.newpage()
grid::grid.draw(p$CD14)
```

```{r}
saveArchRProject(ArchRProj = projHeme2, outputDirectory = "Save-ProjHeme2", load = FALSE)
```


# scRNAseq integration
```{r}
seRNA <- readRDS("scRNA-Hematopoiesis-Granja-2019.rds")
```

## Unconstrained Integration
```{r}
projHeme2 <- addGeneIntegrationMatrix(
    ArchRProj = projHeme2, 
    useMatrix = "GeneScoreMatrix",
    matrixName = "GeneIntegrationMatrix",
    reducedDims = "IterativeLSI",
    seRNA = seRNA,
    addToArrow = FALSE,
    groupRNA = "BioClassification",
    nameCell = "predictedCell_Un",
    nameGroup = "predictedGroup_Un",
    nameScore = "predictedScore_Un"
)
```

```{r}
head(projHeme2$predictedGroup_Un)
```

## Constrained Integration
```{r}
# The predicted cell type for each scATAC-seq cluster
cM <- as.matrix(confusionMatrix(projHeme2$Clusters, projHeme2$predictedGroup_Un))
preClust <- colnames(cM)[apply(cM, 1 , which.max)]
cbind(preClust, rownames(cM)) #Assignments
```

```{r}
table(seRNA@colData$BioClassification)
```

```{r}
#scRNA-seq 19-25 are T or NK cells
cTNK <- paste0(paste0(19:25), collapse="|")
cTNK
```

```{r}
#scRNA-seq other cells
cNonTNK <- paste0(c(paste0("0", 1:9), 10:13, 15:18), collapse="|")
cNonTNK
```

```{r}
#scATAC T or NK clusters
clustTNK <- rownames(cM)[grep(cTNK, preClust)]
clustTNK
```

```{r}
#scATAC none T or NK clusters
clustNonTNK <- rownames(cM)[grep(cNonTNK, preClust)]
clustNonTNK
```

```{r}
#scRNA T NK cell names
rnaTNK <- colnames(seRNA)[grep(cTNK, colData(seRNA)$BioClassification)]
head(rnaTNK)
```

```{r}
rnaNonTNK <- colnames(seRNA)[grep(cNonTNK, colData(seRNA)$BioClassification)]
head(rnaNonTNK)
```

```{r}
groupList <- SimpleList(
    TNK = SimpleList(
        ATAC = projHeme2$cellNames[projHeme2$Clusters %in% clustTNK],
        RNA = rnaTNK
    ),
    NonTNK = SimpleList(
        ATAC = projHeme2$cellNames[projHeme2$Clusters %in% clustNonTNK],
        RNA = rnaNonTNK
    )    
)
```

```{r}
projHeme2 <- addGeneIntegrationMatrix(
    ArchRProj = projHeme2, 
    useMatrix = "GeneScoreMatrix",
    matrixName = "GeneIntegrationMatrix",
    reducedDims = "IterativeLSI",
    seRNA = seRNA,
    addToArrow = FALSE, 
    groupList = groupList,
    groupRNA = "BioClassification",
    nameCell = "predictedCell_Co",
    nameGroup = "predictedGroup_Co",
    nameScore = "predictedScore_Co"
)
```

```{r}
pal <- paletteDiscrete(values = colData(seRNA)$BioClassification)
pal
```

```{r}
p1 <- plotEmbedding(
    projHeme2, 
    colorBy = "cellColData", 
    name = "predictedGroup_Un", 
    pal = pal
)
p2 <- plotEmbedding(
    projHeme2, 
    colorBy = "cellColData", 
    name = "predictedGroup_Co", 
    pal = pal
)
p1+p2
```

```{r}
projHeme3 <- addGeneIntegrationMatrix(
    ArchRProj = projHeme2, 
    useMatrix = "GeneScoreMatrix",
    matrixName = "GeneIntegrationMatrix",
    reducedDims = "IterativeLSI",
    seRNA = seRNA,
    addToArrow = TRUE,
    force= TRUE,
    groupList = groupList,
    groupRNA = "BioClassification",
    nameCell = "predictedCell",
    nameGroup = "predictedGroup",
    nameScore = "predictedScore"
)
```

```{r}
getAvailableMatrices(projHeme3)
```

```{r}
projHeme3 <- addImputeWeights(projHeme3)

```

```{r}
markerGenes  <- c(
    "CD34", #Early Progenitor
    "GATA1", #Erythroid
    "PAX5", "MS4A1", #B-Cell Trajectory
    "CD14", #Monocytes
    "CD3D", "CD8A", "TBX21", "IL7R" #TCells
  )

p1 <- plotEmbedding(
    ArchRProj = projHeme3, 
    colorBy = "GeneIntegrationMatrix", 
    name = markerGenes, 
    continuousSet = "horizonExtra",
    embedding = "UMAP",
    imputeWeights = getImputeWeights(projHeme3)
)

p2 <- plotEmbedding(
    ArchRProj = projHeme3, 
    colorBy = "GeneScoreMatrix", 
    continuousSet = "horizonExtra",
    name = markerGenes, 
    embedding = "UMAP",
    imputeWeights = getImputeWeights(projHeme3)
)
```

```{r}
p1c <- lapply(p1, function(x){
    x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
        axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()
    )
})

p2c <- lapply(p2, function(x){
    x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
        axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()
    )
})

# Gene expression deduced by scRNA-seq
do.call(cowplot::plot_grid, c(list(ncol = 3), p1c))
```
```{r}
#Gene score
do.call(cowplot::plot_grid, c(list(ncol = 3), p2c))

```

```{r}
cM <- confusionMatrix(projHeme3$Clusters, projHeme3$predictedGroup)
labelOld <- rownames(cM)
labelOld
```

```{r}
labelNew <- colnames(cM)[apply(cM, 1, which.max)]
labelNew
```

```{r}
remapClust <- c(
    "01_HSC" = "Progenitor",
    "02_Early.Eryth" = "Erythroid",
    "03_Late.Eryth" = "Erythroid",
    "04_Early.Baso" = "Basophil",
    "05_CMP.LMPP" = "Progenitor",
    "06_CLP.1" = "CLP",
    "07_GMP" = "GMP",
    "08_GMP.Neut" = "GMP",
    "09_pDC" = "pDC",
    "10_cDC" = "cDC",
    "11_CD14.Mono.1" = "Mono",
    "12_CD14.Mono.2" = "Mono",
    "13_CD16.Mono" = "Mono",
    "15_CLP.2" = "CLP",
    "16_Pre.B" = "PreB",
    "17_B" = "B",
    "18_Plasma" = "Plasma",
    "19_CD8.N" = "CD8.N",
    "20_CD4.N1" = "CD4.N",
    "21_CD4.N2" = "CD4.N",
    "22_CD4.M" = "CD4.M",
    "23_CD8.EM" = "CD8.EM",
    "24_CD8.CM" = "CD8.CM",
    "25_NK" = "NK"
)
remapClust <- remapClust[names(remapClust) %in% labelNew]
```

```{r}
labelNew2 <- mapLabels(labelNew, oldLabels = names(remapClust), newLabels = remapClust)
labelNew2
```

```{r}
projHeme3$Clusters2 <- mapLabels(projHeme3$Clusters, newLabels = labelNew2, oldLabels = labelOld)

```

```{r}
plotEmbedding(projHeme3, colorBy = "cellColData", name = "Clusters2")
```

```{r}
saveArchRProject(ArchRProj = projHeme3, outputDirectory = "Save-ProjHeme3", load = FALSE)

```

# Pseudo-bulk replicates

```{r}
library(BSgenome.Hsapiens.UCSC.hg19, lib.loc = "/Library/Frameworks/R.framework/Versions/4.2/Resources/library")
projHeme4 <- addGroupCoverages(ArchRProj = projHeme3, groupBy = "Clusters2")

```

# Peak calling
```{r}
projHeme4 <- addReproduciblePeakSet(
    ArchRProj = projHeme4, 
    groupBy = "Clusters2", 
    pathToMacs2 = "/Users/nzhang/opt/anaconda3/envs/py37/bin/macs2"
)
```

```{r}
getPeakSet(projHeme4)

```


```{r}
saveArchRProject(ArchRProj = projHeme4, outputDirectory = "Save-ProjHeme4", load = FALSE)
```

```{r}
projHeme5 <- addPeakMatrix(projHeme4)

```

```{r}
getAvailableMatrices(projHeme5)
```

```{r}
markersPeaks <- getMarkerFeatures(
    ArchRProj = projHeme5, 
    useMatrix = "PeakMatrix", 
    groupBy = "Clusters2",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  testMethod = "wilcoxon"
)
```

```{r}
markersPeaks
```

```{r}
markerList <- getMarkers(markersPeaks, cutOff = "FDR <= 0.01 & Log2FC >= 1", returnGR = TRUE)
markerList
```

```{r}
markerList$B
```

```{r}
heatmapPeaks <- markerHeatmap(
  seMarker = markersPeaks, 
  cutOff = "FDR <= 0.1 & Log2FC >= 0.5",
  transpose = TRUE
)
```


```{r}
draw(heatmapPeaks, heatmap_legend_side = "bot", annotation_legend_side = "bot")

```

```{r}
pv <- markerPlot(seMarker = markersPeaks, name = "Erythroid", cutOff = "FDR <= 0.1 & Log2FC >= 1", plotAs = "Volcano")
pv
```

```{r}
p <- plotBrowserTrack(
    ArchRProj = projHeme5, 
    groupBy = "Clusters2", 
    geneSymbol = c("GATA1"),
    features =  getMarkers(markersPeaks, cutOff = "FDR <= 0.1 & Log2FC >= 1", returnGR = TRUE)["Erythroid"],
    upstream = 50000,
    downstream = 50000
)
```
```{r}
grid::grid.newpage()
grid::grid.draw(p$GATA1)
```

```{r}
markerTest <- getMarkerFeatures(
  ArchRProj = projHeme5, 
  useMatrix = "PeakMatrix",
  groupBy = "Clusters2",
  testMethod = "wilcoxon",
  bias = c("TSSEnrichment", "log10(nFrags)"),
  useGroups = "Erythroid",
  bgdGroups = "Progenitor"
)
```

```{r}
markerPlot(seMarker = markerTest, name = "Erythroid", cutOff = "FDR <= 0.1 & abs(Log2FC) >= 1", plotAs = "Volcano")
```
```{r}
saveArchRProject(ArchRProj = projHeme5, outputDirectory = "Save-ProjHeme5", load = FALSE)
```

# Motif and Feature Enrichment

```{r}
projHeme5 <- addMotifAnnotations(ArchRProj = projHeme5, motifSet = "cisbp", name = "Motif")
```

```{r}
motifsUp <- peakAnnoEnrichment(
    seMarker = markerTest,
    ArchRProj = projHeme5,
    peakAnnotation = "Motif",
    cutOff = "FDR <= 0.1 & Log2FC >= 0.5"
  )
```

```{r}
motifsUp
```

```{r}
df <- data.frame(TF = rownames(motifsUp), mlog10Padj = assay(motifsUp)[,1])
df <- df[order(df$mlog10Padj, decreasing = TRUE),]
df$rank <- seq_len(nrow(df))
```

```{r}
head(df)
```

```{r}
ggplot(df, aes(rank, mlog10Padj, color = mlog10Padj)) + 
  geom_point(size = 1) +
  ggrepel::geom_label_repel(
        data = df[rev(seq_len(30)), ], aes(x = rank, y = mlog10Padj, label = TF), 
        size = 1.5,
        nudge_x = 2,
        color = "black"
  ) + theme_ArchR() + 
  ylab("-log10(P-adj) Motif Enrichment") + 
  xlab("Rank Sorted TFs Enriched") +
  scale_color_gradientn(colors = paletteContinuous(set = "comet"))
```
```{r}
enrichMotifs <- peakAnnoEnrichment(
    seMarker = markersPeaks,
    ArchRProj = projHeme5,
    peakAnnotation = "Motif",
    cutOff = "FDR <= 0.1 & Log2FC >= 0.5"
  )
```

```{r}
heatmapEM <- plotEnrichHeatmap(enrichMotifs, n = 7, transpose = TRUE)
ComplexHeatmap::draw(heatmapEM, heatmap_legend_side = "bot", annotation_legend_side = "bot")

```
## The overlap with bulk ATAC-seq peak set
```{r}
projHeme5 <- addArchRAnnotations(ArchRProj = projHeme5, collection = "ATAC")
enrichATAC <- peakAnnoEnrichment(
    seMarker = markersPeaks,
    ArchRProj = projHeme5,
    peakAnnotation = "ATAC",
    cutOff = "FDR <= 0.1 & Log2FC >= 0.5"
  )
```

```{r}
heatmapATAC <- plotEnrichHeatmap(enrichATAC, n = 7, transpose = TRUE)
ComplexHeatmap::draw(heatmapATAC, heatmap_legend_side = "bot", annotation_legend_side = "bot")

```

# ChromVAR Deviatons Enrichment
## chromVAR predicts enrichment of TF activity on a per-cell basis. It also corrects the Tn5 transposase insertion sequence bias.
```{r}
if("Motif" %ni% names(projHeme5@peakAnnotation)){
    projHeme5 <- addMotifAnnotations(ArchRProj = projHeme5, motifSet = "cisbp", name = "Motif")
}
```

```{r}
projHeme5 <- addBgdPeaks(projHeme5)

```

```{r}
# This function save the result into the "MotifMatrix" matrix which can be used later.
projHeme5 <- addDeviationsMatrix(
  ArchRProj = projHeme5, 
  peakAnnotation = "Motif",
  force = TRUE
)
```

## The variation of each TF motif across all cells.
```{r}
plotVarDev <- getVarDeviations(projHeme5, name = "MotifMatrix", plot = TRUE)
```


```{r}
plotVarDev
```
## Check the variation of target TF motifs across all cells.
### Get the motif names corresponding to each TF
```{r}
motifs <- c("GATA1", "CEBPA", "EBF1", "IRF4", "TBX21", "PAX5")
markerMotifs <- getFeatures(projHeme5, select = paste(motifs, collapse="|"), useMatrix = "MotifMatrix")
markerMotifs
```

### Motif names clean up
```{r}
markerMotifs <- grep("z:", markerMotifs, value = TRUE)
markerMotifs <- markerMotifs[markerMotifs %ni% "z:SREBF1_22"]
markerMotifs
```

```{r}
projHeme5 <- addImputeWeights(projHeme5)
```

```{r}
p <- plotGroups(ArchRProj = projHeme5, 
  groupBy = "Clusters2", 
  colorBy = "MotifMatrix", 
  name = markerMotifs,
  imputeWeights = getImputeWeights(projHeme5)
)
```

```{r}
p2 <- lapply(seq_along(p), function(x){
  if(x != 1){
    p[[x]] + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6) +
    theme(plot.margin = unit(c(0.1, 0.1, 0.1, 0.1), "cm")) +
    theme(
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank(),
        axis.title.y=element_blank()
    ) + ylab("")
  }else{
    p[[x]] + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6) +
    theme(plot.margin = unit(c(0.1, 0.1, 0.1, 0.1), "cm")) +
    theme(
        axis.ticks.y=element_blank(),
        axis.title.y=element_blank()
    ) + ylab("")
  }
})
do.call(cowplot::plot_grid, c(list(nrow = 1, rel_widths = c(2, rep(1, length(p2) - 1))),p2))
```
### Plot the motif enrichment z-scores on UMAP
```{r}
p <- plotEmbedding(
    ArchRProj = projHeme5, 
    colorBy = "MotifMatrix", 
    name = sort(markerMotifs), 
    embedding = "UMAP",
    imputeWeights = getImputeWeights(projHeme5)
)
```

```{r}
p2 <- lapply(p, function(x){
    x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
        axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()
    )
})
do.call(cowplot::plot_grid, c(list(ncol = 3),p2))
```
### Plot the previously infered gene expression scores on UMAP and compare with motif enrichment z-scores
```{r}
markerRNA <- getFeatures(projHeme5, select = paste(motifs, collapse="|"), useMatrix = "GeneScoreMatrix")
markerRNA <- markerRNA[markerRNA %ni% c("SREBF1","CEBPA-DT")]
markerRNA
```

```{r}
p <- plotEmbedding(
    ArchRProj = projHeme5, 
    colorBy = "GeneScoreMatrix", 
    name = sort(markerRNA), 
    embedding = "UMAP",
    imputeWeights = getImputeWeights(projHeme5)
)
```

```{r}
p2 <- lapply(p, function(x){
    x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
        axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()
    )
})
do.call(cowplot::plot_grid, c(list(ncol = 3),p2))
```

### We previously linked our scATAC-seq data with corresponding scRNA-seq data, also plot the linked gene expression values

```{r}
markerRNA <- getFeatures(projHeme5, select = paste(motifs, collapse="|"), useMatrix = "GeneIntegrationMatrix")
markerRNA <- markerRNA[markerRNA %ni% c("SREBF1","CEBPA-DT")]
markerRNA
```

```{r}
p <- plotEmbedding(
    ArchRProj = projHeme5, 
    colorBy = "GeneIntegrationMatrix", 
    name = sort(markerRNA), 
    embedding = "UMAP",
    continuousSet = "blueYellow",
    imputeWeights = getImputeWeights(projHeme5)
)
```

```{r}
p2 <- lapply(p, function(x){
    x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
        axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()
    )
})
do.call(cowplot::plot_grid, c(list(ncol = 3),p2))
```
## ChromVAR custom deviation
### Calculate enrichment using bulk ATAC-seq peaks
```{r}
if("ATAC" %ni% names(projHeme5@peakAnnotation)){
    projHeme5 <- addArchRAnnotations(ArchRProj = projHeme5, collection = "ATAC")
}
```

```{r}
projHeme5 <- addDeviationsMatrix(
  ArchRProj = projHeme5, 
  peakAnnotation = "ATAC",
  force = TRUE
)
```

```{r}
plotVarDev <- getVarDeviations(projHeme5, plot = TRUE, name = "ATACMatrix")
plotVarDev
```

```{r}
ATACPeaks <- c("Heme_HSC", "Heme_LMPP", "Heme_Ery", "Heme_Mono", "Heme_CD4", "Heme_CD8", "Heme_B", "Heme_NK", "IAtlas_DC_Plasmacytoid")
markerATAC <- getFeatures(projHeme5, select = paste(ATACPeaks, collapse="|"), useMatrix = "ATACMatrix")
markerATAC <- sort(grep("z:", markerATAC, value = TRUE))
markerATAC
```

```{r}
p <- plotEmbedding(
    ArchRProj = projHeme5, 
    colorBy = "ATACMatrix", 
    name = markerATAC, 
    embedding = "UMAP",
    imputeWeights = getImputeWeights(projHeme5)
)
```

```{r}
p2 <- lapply(p, function(x){
    x + guides(color = FALSE, fill = FALSE) + 
    theme_ArchR(baseSize = 6.5) +
    theme(plot.margin = unit(c(0, 0, 0, 0), "cm")) +
    theme(
        axis.text.x=element_blank(), 
        axis.ticks.x=element_blank(), 
        axis.text.y=element_blank(), 
        axis.ticks.y=element_blank()
    )
})
do.call(cowplot::plot_grid, c(list(ncol = 3),p2))
```
# Motif footprinting
```{r}
#Get the "Motif" (CIS-BP motifs) from peakAnnotation slot
motifPositions <- getPositions(projHeme5,name="Motif")
```

```{r}
motifPositions
```

```{r}
motifs <- c("GATA1", "CEBPA", "EBF1", "IRF4", "TBX21", "PAX5")
markerMotifs <- unlist(lapply(motifs, function(x) grep(x, names(motifPositions), value = TRUE)))
markerMotifs <- markerMotifs[markerMotifs %ni% "SREBF1_22"]
markerMotifs
```


```{r}
seFoot <- getFootprints(
  ArchRProj = projHeme5, 
  positions = motifPositions[markerMotifs], 
  groupBy = "Clusters2"
)
```

## Normalization of Footprints for Tn5 Bias
### Without Tn5 Bias normalization, Subtracting the Tn5 Bias or Dividing by the Tn5 Bias. Results in /HemeOutput/Plots

```{r}
plotFootprints(
  seFoot = seFoot,
  ArchRProj = projHeme5, 
  normMethod = "None",
  plotName = "Footprints-No-Normalization",
  addDOC = FALSE,
  smoothWindow = 5
)
```

```{r}
plotFootprints(
  seFoot = seFoot,
  ArchRProj = projHeme5, 
  normMethod = "Subtract",
  plotName = "Footprints-Subtract-Bias",
  addDOC = FALSE,
  smoothWindow = 5
)
```

```{r}
plotFootprints(
  seFoot = seFoot,
  ArchRProj = projHeme5, 
  normMethod = "Divide",
  plotName = "Footprints-Divide-Bias",
  addDOC = FALSE,
  smoothWindow = 5
)
```

### Footprinting with user-defined feature set
```{r}
seTSS <- getFootprints(
  ArchRProj = projHeme5, 
  positions = GRangesList(TSS = getTSS(projHeme5)), 
  groupBy = "Clusters2",
  flank = 2000
)

plotFootprints(
  seFoot = seTSS,
  ArchRProj = projHeme5, 
  normMethod = "None",
  plotName = "TSS-No-Normalization",
  addDOC = FALSE,
  flank = 2000,
  flankNorm = 100
)
```

# Co-accessibility

```{r}
projHeme5 <- addCoAccessibility(
    ArchRProj = projHeme5,
    reducedDims = "IterativeLSI"
)
```

```{r}
cA <- getCoAccessibility(
    ArchRProj = projHeme5,
    corCutOff = 0.5,
    resolution = 1,
    returnLoops = FALSE
)
```

```{r}
cA
```

```{r}
metadata(cA)[[1]]
```


```{r}
metadata(cA)[[1]][5]
```

```{r}
metadata(cA)[[1]][8]
```

## Plot co-accessibility browser tracks
```{r}
markerGenes  <- c(
    "CD34", #Early Progenitor
    "GATA1", #Erythroid
    "PAX5", "MS4A1", #B-Cell Trajectory
    "CD14", #Monocytes
    "CD3D", "CD8A", "TBX21", "IL7R" #TCells
  )

p <- plotBrowserTrack(
    ArchRProj = projHeme5, 
    groupBy = "Clusters2", 
    geneSymbol = markerGenes, 
    upstream = 50000,
    downstream = 50000,
    loops = getCoAccessibility(projHeme5)
)
```

```{r}
grid::grid.newpage()
grid::grid.draw(p$CD14)
```

# Peak2GeneLinkage
## Correlate peak accessibility to scRNA-seq gene expression

```{r}
projHeme5 <- addPeak2GeneLinks(
    ArchRProj = projHeme5,
    reducedDims = "IterativeLSI"
)
```

```{r}
p2g <- getPeak2GeneLinks(
    ArchRProj = projHeme5,
    corCutOff = 0.45,
    resolution = 1,
    returnLoops = FALSE
)
```


```{r}
p2g
```

```{r}
metadata(p2g)[[1]]
```

```{r}
metadata(p2g)
```

## Plot peak-to-gene links
```{r}
p <- plotBrowserTrack(
    ArchRProj = projHeme5, 
    groupBy = "Clusters2", 
    geneSymbol = markerGenes, 
    upstream = 50000,
    downstream = 50000,
    loops = getPeak2GeneLinks(projHeme5)
)
```

```{r}
grid::grid.newpage()
grid::grid.draw(p$CD14)
```

## Heatmap of peak-to-gene links
### Each row is one peak2gene link, each column is one cell
```{r}
p <- plotPeak2GeneHeatmap(ArchRProj = projHeme5, groupBy = "Clusters2")
p
```

# Identify TF regulators by combining chromVAR with scRNA-seq
# Identify the TF whose gene expression is closely correlated with its motif enrichment (MotifMatrix) and its motif enrichment is significantly different across cell clusters.
## MotifMatrix of chromVAR deviations average by clusters
```{r}
seGroupMotif <- getGroupSE(ArchRProj = projHeme5, useMatrix = "MotifMatrix", groupBy = "Clusters2")
seGroupMotif
```

### seGroupMotif has two seqnames: 1. "deviations": raw deviation. 2. "z": deviation z-score.
### Only use the z scores here.
```{r}
seZ <- seGroupMotif[rowData(seGroupMotif)$seqnames=="z",]
seZ
```

```{r}
head(assay(seZ))
```

```{r}
head(rowData(seZ))
```

```{r}
head(colData(seZ))
```

### Identify the maximum delta in z-score between all clusters
```{r}
rowData(seZ)$maxDelta <- lapply(seq_len(ncol(seZ)), function(x){
  rowMaxs(assay(seZ) - assay(seZ)[,x])
}) %>% Reduce("cbind", .) %>% rowMaxs
```

```{r}
head(rowData(seZ))
```

## Calculate the correlation between MotifMatrix and Gene score/ integrated expression
```{r}
corGSM_MM <- correlateMatrices(
    ArchRProj = projHeme5,
    useMatrix1 = "GeneScoreMatrix",
    useMatrix2 = "MotifMatrix",
    reducedDims = "IterativeLSI"
)
```

```{r}
head(corGSM_MM)
```

```{r}
corGIM_MM <- correlateMatrices(
    ArchRProj = projHeme5,
    useMatrix1 = "GeneIntegrationMatrix",
    useMatrix2 = "MotifMatrix",
    reducedDims = "IterativeLSI"
)
```

```{r}
head(corGIM_MM)
```

## Add MotifMatrix max delta to the correlation data frame

```{r}
corGSM_MM$maxDelta <- rowData(seZ)[match(corGSM_MM$MotifMatrix_name, rowData(seZ)$name), "maxDelta"]
corGIM_MM$maxDelta <- rowData(seZ)[match(corGIM_MM$MotifMatrix_name, rowData(seZ)$name), "maxDelta"]
```

## Identify positive TF regulators
```{r}
corGSM_MM <- corGSM_MM[order(abs(corGSM_MM$cor), decreasing = TRUE), ]
corGSM_MM <- corGSM_MM[which(!duplicated(gsub("\\-.*","",corGSM_MM[,"MotifMatrix_name"]))), ]
corGSM_MM$TFRegulator <- "NO"
corGSM_MM$TFRegulator[which(corGSM_MM$cor > 0.5 & corGSM_MM$padj < 0.01 & corGSM_MM$maxDelta > quantile(corGSM_MM$maxDelta, 0.75))] <- "YES"
sort(corGSM_MM[corGSM_MM$TFRegulator=="YES",1])
```

```{r}
ggplot(data.frame(corGSM_MM), aes(cor, maxDelta, color = TFRegulator)) +
  geom_point() + 
  theme_ArchR() +
  geom_vline(xintercept = 0, lty = "dashed") + 
  scale_color_manual(values = c("NO"="darkgrey", "YES"="firebrick3")) +
  xlab("Correlation To Gene Score") +
  ylab("Max TF Motif Delta") +
  scale_y_continuous(
    expand = c(0,0), 
    limits = c(0, max(corGSM_MM$maxDelta)*1.05)
  )
```
```{r}
corGIM_MM <- corGIM_MM[order(abs(corGIM_MM$cor), decreasing = TRUE), ]
corGIM_MM <- corGIM_MM[which(!duplicated(gsub("\\-.*","",corGIM_MM[,"MotifMatrix_name"]))), ]
corGIM_MM$TFRegulator <- "NO"
corGIM_MM$TFRegulator[which(corGIM_MM$cor > 0.5 & corGIM_MM$padj < 0.01 & corGIM_MM$maxDelta > quantile(corGIM_MM$maxDelta, 0.75))] <- "YES"
sort(corGIM_MM[corGIM_MM$TFRegulator=="YES",1])
```

```{r}
ggplot(data.frame(corGIM_MM), aes(cor, maxDelta, color = TFRegulator)) +
  geom_point() + 
  theme_ArchR() +
  geom_vline(xintercept = 0, lty = "dashed") + 
  scale_color_manual(values = c("NO"="darkgrey", "YES"="firebrick3")) +
  xlab("Correlation To Gene Expression") +
  ylab("Max TF Motif Delta") +
  scale_y_continuous(
    expand = c(0,0), 
    limits = c(0, max(corGIM_MM$maxDelta)*1.05)
  )
```

# Trajectory analysis
```{r}
p1 <- plotEmbedding(ArchRProj = projHeme5, colorBy = "cellColData", name = "Clusters", embedding = "UMAP")
p2 <- plotEmbedding(ArchRProj = projHeme5, colorBy = "cellColData", name = "Clusters2", embedding = "UMAP")
ggAlignPlots(p1, p2, type = "h")
```
## Myeloid Trajectory - Monocyte Differentiation
```{r}
trajectory <- c("Progenitor", "GMP", "Mono")
projHeme5 <- addTrajectory(
    ArchRProj = projHeme5, 
    name = "MyeloidU", 
    groupBy = "Clusters2",
    trajectory = trajectory, 
    embedding = "UMAP", 
    force = TRUE
)
```

### Each cell has a unique pseudotime value between 0 and 100. 
```{r}
head(projHeme5$MyeloidU[!is.na(projHeme5$MyeloidU)])
```

```{r}
p=plotTrajectory(projHeme5, trajectory = "MyeloidU", colorBy = "cellColData", name = "MyeloidU")
p[[1]]
```

### Plot features along the trajectory
```{r}
projHeme5 <- addImputeWeights(projHeme5)
```

```{r}
p1 <- plotTrajectory(projHeme5, trajectory = "MyeloidU", colorBy = "GeneScoreMatrix", name = "CEBPB", continuousSet = "horizonExtra")
p2 <- plotTrajectory(projHeme5, trajectory = "MyeloidU", colorBy = "GeneIntegrationMatrix", name = "CEBPB", continuousSet = "blueYellow")
ggAlignPlots(p1[[1]], p2[[1]], type = "h")
```

```{r}
ggAlignPlots(p1[[2]], p2[[2]], type = "h")
```
### Plot several features along pseudo-time using heatmaps
```{r}
trajMM  <- getTrajectory(ArchRProj = projHeme5, name = "MyeloidU", useMatrix = "MotifMatrix", log2Norm = FALSE)
```

```{r}
plotTrajectoryHeatmap(trajMM, pal = paletteContinuous(set = "solarExtra"))
```
```{r}
trajGSM <- getTrajectory(ArchRProj = projHeme5, name = "MyeloidU", useMatrix = "GeneScoreMatrix", log2Norm = TRUE)
```

```{r}
trajectoryHeatmap(trajGSM,  pal = paletteContinuous(set = "horizonExtra"))
```
### Correlation between features along pseudo-time
```{r}
corGSM_MM <- correlateTrajectories(trajGSM, trajMM)
```

```{r}
corGSM_MM[[1]]
```

```{r}
trajGSM2 <- trajGSM[corGSM_MM[[1]]$name1, ]
trajMM2 <- trajMM[corGSM_MM[[1]]$name2, ]
```

```{r}
trajCombined <- trajGSM2
assay(trajCombined, withDimnames=FALSE) <- t(apply(assay(trajGSM2), 1, scale)) + t(apply(assay(trajMM2), 1, scale))
```

```{r}
combinedMat <- plotTrajectoryHeatmap(trajCombined, returnMat = TRUE, varCutOff = 0)
```

```{r}
rowOrder <- match(rownames(combinedMat), rownames(trajGSM2))
```

```{r}
ht1 <- plotTrajectoryHeatmap(trajGSM2,  pal = paletteContinuous(set = "horizonExtra"),  varCutOff = 0, rowOrder = rowOrder)
```

```{r}
ht2 <- plotTrajectoryHeatmap(trajMM2, pal = paletteContinuous(set = "solarExtra"), varCutOff = 0, rowOrder = rowOrder)
```

```{r}
ht1 + ht2
```



